---
title: "Big Data Bowl: Endzone Kickoff Exploration"
output: 
  html_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.height = 8, fig.width = 12, message = F, warning = F)

library(tidyverse)
library(lubridate)
library(stringr)
library(reactable)
library(highcharter)
library(purrr)
library(here)
library(feather)
library(skimr)
library(nflfastR)
library(ggimage)
library(RCurl)
library(here)

source(here::here("R", "util.R"))

# load data

# # read in analysis data files
# file_names <- 
#   c("kickoff_all", "kickoffs", "pbp_kickoff", "scouting_kickoff", "tracking_kickoff", "tracking_kickoff_ball")
# 
# list.files(path = here::here("data", "analysis"), pattern = ".feather") %>%
#   map(., ~load_data(type = "analysis", file_name = .x)) %>% 
#   set_names(nm = file_names) %>% 
#   list2env(., envir = .GlobalEnv)

file_names <- 
  c("kickoff_all", "kickoffs", "kickoff_tracking_avg_speed", "pbp_kickoff", "scouting_kickoff", "tracking_kickoff", "tracking_kickoff_ball")

list.files(path = here::here("data", "analysis"), pattern = ".feather") %>%
  map(., ~load_data(type = "analysis", file_name = .x)) %>% 
  set_names(nm = file_names) %>% 
  list2env(., envir = .GlobalEnv)

# remove(tracking_kickoff_ball, scouting_kickoff, kickoff_all, tracking_kickoff)

# team logos
url_logo <- 
  getURL("https://raw.githubusercontent.com/statsbylopez/BlogPosts/master/nfl_teamlogos.csv")
df_logos <- 
  read.csv(text = url_logo)
```

<style type="text/css">
.main-container {
  max-width: 1200px;
  margin-left: auto;
  margin-right: auto;
}
</style>

## Intro
The goal of this analysis is to accurately predict when a receiving team will return a kick out of the endzone *instead* of taking a touchback. We will explore some potential differences between kickoffs that are returned out of the endzone vs those that result in a touchback.

## Data Exploration

After eliminating kickoffs that did not reach the endzone, we can start to explore the data

### Kickoffs

```{r}
kickoff_bar <- 
  kickoffs %>%
  filter(return_type %in% c("Endzone Return", "Touchback")) %>%
  group_by(season, return_type) %>%
  tally() %>%
  mutate(freq = n/ sum(n)) %>%
  ggplot(., aes(x = return_type, y = n, fill = return_type, label = paste(round(freq*100, 2), "%"))) + 
  geom_col(alpha = .75) +
  geom_text(nudge_y = 20) + 
  theme_minimal() + 
  theme(legend.position = 'none',
        text = add_big_labels(text_size = 16)) +
  facet_wrap(.~season) + 
  scale_fill_manual(values = c("cornflowerblue", "tomato")) + 
  labs(x = "", y = "Count")

# title = "Endzone returns are somewhat consistent across seasons",
#        subtitle = "About 19% of kickoffs that rach the endzone are returned out of the endzone", align = 'left')


ggsave(kickoff_bar, filename = here::here("img", "kickoff_bar.svg"))
```

```{r}
team_stacked <- 
  kickoffs %>%
  filter(return_type %in% c("Endzone Return", "Touchback")) %>%
  group_by(recieving_team_name, return_type) %>%
  tally() %>%
  mutate(freq = n/ sum(n)) %>%
  ungroup() %>%
  group_by(recieving_team_name) %>%
  mutate(return_value = ifelse(return_type == "Endzone Return", freq, NA_real_),
         return_value = max(return_value, na.rm = T)) %>%
  ungroup() %>%
  arrange(desc(return_value)) 

ggplot(team_stacked, aes(x = reorder(recieving_team_name, return_value), y = freq, fill = as.character(return_type), label = paste0(round(freq*100, 2), "%"))) + 
  geom_col(position = "stack", alpha = .75) +
  geom_text(position = position_stack(vjust = .5), show.legend = F) +
  coord_flip() +
  theme_minimal() + 
  theme(text = add_big_labels(text_size = 16)) +
  # facet_wrap(.~season) +
  scale_fill_manual(values = c("cornflowerblue", "tomato")) +
  scale_y_continuous(expand = c(0, .01), labels = scales::percent)  +
  labs(x = "", y = "", title = "Endzone return types vary by team", align = 'left', fill = NULL)
```

```{r}
# starting yardline vs endzone kick depth
kickoffs %>%
  filter(return_type == "Endzone Return") %>%
  group_by(game_id, recieving_team_name) %>%
  summarise(n = n(),
            median_starting_yardline = median(starting_yardline, na.rm = T)) %>%
  ungroup() %>%
  group_by(recieving_team_name) %>%
  summarise(mean_n = mean(n),
         median_starting_yardline = median(median_starting_yardline)) %>%
  ungroup() %>%
  mutate(overall_mean_n = mean(mean_n),
         overall_median_yards = median(median_starting_yardline, na.rm = T)) %>%
  left_join(., df_logos, by = c("recieving_team_name" = "team_code")) %>%
  ggplot(., aes(x = median_starting_yardline, y = mean_n)) +
  geom_image(aes(image = url), size = 0.04) + 
  geom_hline(aes(yintercept = overall_mean_n), linetype = 'dashed') +
  annotate("text", x = 25, y = 1.45, label = "Median: 1.49", vjust = .5) +
  geom_vline(aes(xintercept = overall_median_yards), linetype = 'dashed')  +
  annotate("text", x = 22.5, y = 2, label = "Mean: 22.75", vjust = -0.5) +
  theme_minimal() +
  theme(text = add_big_labels(text_size = 16)) +
  theme(legend.position = 'none') +
  labs(y = "Endzone Returns per Game\n", x = "Starting Field Position",
       title = "Endzone Kickoff Returns")
```

```{r}
# Starting yardline boxplot
yardline <- 
  kickoffs %>%
  filter(return_type != "Touchback") %>%
  ggplot(., aes(x = return_type, y = starting_yardline, fill = return_type)) + 
  geom_jitter(alpha = .3, color = 'gray') + 
  geom_boxplot(alpha = .75, outlier.shape = NA) + 
  coord_flip() + 
  geom_hline(yintercept = 25, linetype = 'dashed', color = "black") +
  scale_y_continuous(breaks= seq(0, 100, 5)) + 
  theme_minimal() + 
  scale_fill_manual(values = c("cornflowerblue", "tomato")) + 
  theme(legend.position = 'none',
        panel.grid.minor = element_blank(),
        text = add_big_labels(text_size = 12)) + 
  labs(x = "", y = "Drive Starting Yardline")

# return type vs EPA
epa <- 
  kickoffs %>%
  left_join(., pbp_kickoff %>% 
              select(old_game_id, play_id, epa) %>% 
              distinct(), by = c("game_id" = "old_game_id", "play_id")) %>%
  filter(return_type != "Field Return",
         between(epa, -4, 4)
         ) %>% 
  ggplot(., aes(x = return_type, y = epa, fill = return_type)) +
  geom_jitter(alpha = .3, color = 'gray') + 
  geom_boxplot(alpha = .75, outlier.shape = NA) + 
  scale_y_continuous(breaks = seq(-4, 4, .5)) +
  coord_flip() +
  theme_minimal() +
  theme(text = add_big_labels(text_size = 12)) + 
  theme(legend.position = 'none') +
  scale_fill_manual(values = c("cornflowerblue", "tomato")) + 
  labs(x = "", y = "EPA")

ggpubr::ggarrange(yardline, epa, ncol = 2, nrow = 1)

ggsave(yardline, filename = here::here("img", "yardline.svg"))
ggsave(epa, filename = here::here("img", "epa.svg"))

```

### Potential Predictors

```{r}

kickoff_with_pbp <- 
  kickoffs %>%
  filter(return_type %in% c("Endzone Return", "Touchback")) %>%
  left_join(., pbp_kickoff, c("game_id" = "old_game_id", "play_id" = "play_id")) %>% 
  mutate(recieving_team_timeouts_remaining = ifelse(recieving_team_name == home_team_abbr, home_timeouts_remaining, away_timeouts_remaining)) %>%
  select(game_id, play_id, return_type, everything())

numeric_cols <- 
  kickoff_with_pbp %>% 
  select(-c(game_id)) %>%
  select(where(is.numeric)) %>%
  names

numeric_boxplots <-
  function(col_name){
    
    # col_name <- "recieving_team_score_diff"
      
    kickoff_with_pbp %>%
      select(return_type, one_of({{col_name}})) %>%
      rename(predictor = 2) %>%
      ggplot(., aes(x = return_type, y = predictor, fill = return_type)) + 
      geom_jitter(alpha = .3, color = 'gray') + 
      geom_boxplot(alpha = .75, outlier.shape = NA) + 
      coord_flip() + 
      theme_minimal() + 
      scale_fill_manual(values = c("cornflowerblue", "tomato")) + 
      labs(x = "", y = "", title = paste("Return type vs", {{col_name}}))
  }

categorical_cols <- 
  kickoff_with_pbp %>% 
  select(-return_type) %>%
  select(where(is.factor) | where(is.character)) %>%
  names

categorical_bar_plots <-
  function(col_name){
    
    kickoff_with_pbp %>%
      select(return_type, one_of({{col_name}})) %>%
      rename(predictor = 2) %>%
      group_by(predictor, return_type) %>%
      tally() %>%
      mutate(freq = n / sum(n)) %>%
      ggplot(., aes(x = reorder(predictor, -freq), y = freq, fill = return_type, label = paste0(paste0(round(freq*100, 2), "%"), "\n(n = ", n, ")"))) + 
      geom_col(position = 'dodge', alpha = .75) + 
      geom_text(position = position_dodge(width = .9), vjust = -0.5) + 
      theme_minimal() + 
      scale_y_continuous(labels = scales::percent, breaks = seq(0, 1, .10)) + 
      scale_fill_manual(values = c("cornflowerblue", "tomato")) + 
      labs(x = "", y = "Count", title = "80% of Endzone Kickoffs Result in a Touchback") + 
      labs(x = "", y = "", title = paste("Return type vs", {{col_name}}))
  }
```

#### Numeric
```{r}
map(numeric_cols, ~numeric_boxplots(.x))
```

#### Categorical
```{r}
map(categorical_cols, ~categorical_bar_plots(.x))
```