---
title: "SPARK: A model to predict kickoff returns"
output: html_document
---

```{r include = FALSE, warning = FALSE, message = FALSE}

library(caret)
library(dplyr)
library(ggplot2)
```
# Introduction

If you ever watched an NFL game, you almost surely have heard an commentator say something like "Man, they could really use a spark right about now". This usually in reference to the losing team, who is in need of an explosive play to gain some momentum and get back in the game. One of the "sparks" that can ignite a team is an incredible kickoff return to give the receiving team great field position, setting the drive up for a higher likelihood of resulting in a field goal or touchdown.

The SPARK model (Special teams Planning A Return of Kickoff) predicts when the receiving team will return a kickoff out of their own endzone *instead* of taking a touchback. 

But why is such a model beneficial? When the receiving team returns a kickoff from their own endzone, they are usually worse off than if they performed a touchback. In turn, a receiving team returning a kickoff from their own endzone *benefits* the kicking team. Returning a kickoff from the endzone, instead of taking a touchback, results in several disadvantages for the receiving team. If the kicking team could better anticipate an endzone return occurring, they may be better off kicking off into the receiving team's endzone *instead* of kicking it short of the goal line.

This analysis uses the data provided, supplemented with play-by-play data from the `nflfastR` package, to develop a model capable of predicting if the receiving team will return a kickoff out of their own endzone with 89% accuracy.

# **S**pecial teams **P**lanning **A** **R**eturn of **K**ickoff (SPARK)

## Motivation

When a kickoff reaches in the endzone, the receiving team has 2 choices: 

 * perform a touchback which removes any chance for a penalty or a turnover, *and* results in 25 free yards, or
 * take their chances running the ball out of the endzone and risk a turnover, a setback from a penalty and/or the possibility of not even reaching the 25 yard line
 
It seems like the best alternative would be to take the touchback, and yet about 20% of kickoffs that reach the endzone are returned out of the endzone by the receiving team! 

**Endzone returns are somewhat consistent across seasons**
![](https://raw.githubusercontent.com/brndngrhm/big-data-bowl/main/img/kickoff_bar.svg)

**Returning a kickoff from the endzone, instead of taking a touchback, results in several disadvantages for the receiving team. Including a worse starting field position (vs both touchback and a field return), and lower EPA**

![](https://raw.githubusercontent.com/brndngrhm/big-data-bowl/main/img/yardline.svg)

![](https://raw.githubusercontent.com/brndngrhm/big-data-bowl/main/img/epa.svg)

## Data Preparation & Feature Engineering

In thinking about what game conditions might lead to a team to run a kickoff out of the endzone, I attempt to create some features to help capture the state of the game just prior to the kickoff. in this way, the data represents the information known at the time of kickoff. Examples of thses features include:

 * the cumulative number of previous endzone return attempts
 * rolling average speed of the kicking team (excluding the kicker) while the ball is in the air
 * result of the drive immediately preceding the kickoff
 
In addition to the provided data sets, play-by-play data from the `nflfastR` package is used to develop some features. The only predictors not known prior to kickoff are kickoff-direction and kickoff-depth (i.e. how far into the endzone the ball is kicked), however, these are variables that can be somewhat controlled by the kicking team.

The full set of features used in the model are as follows. Note: kickoffs that did not reach the endzone were eliminated from the analysis data

#### game-level features

 * seasons week number
 * game day of week
 * game month
 * prime time game indicator
 * stadium surface
 * stadium roof
 * divisional game indicator
 
#### play-level features

 * kickoff from yardline
 * quarter
 * kick length
 * previous endzone return attempts
 * seconds remaining in the half
 * rolling average return yards allowed by the kicking team
 * game half
 * game quarter
 * kickoff yardline
 * result of the drive immediately preceding the kickoff
 * indicator for of the drive immediately preceding the kickoff resulted in a lead change
 * kick direction
 * average speed of the kicking team (excluding the kicker) while the ball is in the air
 * number of timeouts the receiving team has remaining
 
## Model Development

A wide variety of classification models were developed and compared, each tuned using bootstrap resampling (stratified on the outcome of return type) and hyperparameters selected based on log loss. The outcome is somewhat unbalanced, since only about 20% of kickoffs that reach the endzone are returned out of the endzone by the receiving team. Therefore in addition to minimizing log loss, a good model must also be able to distinguish between outcomes so ROC-AUC was considered when selecting a model.

The final model chosen was a random forest. Of all models considered, it had the lowest log loss, as well as the highest accuracy and a competitive ROC-AUC value.

![](https://raw.githubusercontent.com/brndngrhm/big-data-bowl/main/img/model_metrics_comparison.svg)

## Results

The random forest "SPARK" model can predict if the receiving team will return a kickoff out of their own endzone with 89% accuracy.

The performance metrics on the test set closely resemble the training metrics, indicating no overfitting is present:
```{r, echo = FALSE}
model_perf <- 
        read.csv("https://raw.githubusercontent.com/brndngrhm/big-data-bowl/main/train_test_performance.csv")

ggplot(model_perf, aes(x = .metric, y = value, fill = type)) + 
        geom_col(position = "dodge", alpha = .75) + 
        theme_minimal() + 
        coord_flip() + 
        scale_fill_manual(values = c("dodgerblue", "tomato"))
```

Specifics of the test set:
```{r echo = FALSE}

pred_classes <- 
        read.csv("https://raw.githubusercontent.com/brndngrhm/big-data-bowl/main/pred_classes.csv") %>%
        mutate(return_type = factor(return_type),
               .pred_class = factor(.pred_class))

confusionMatrix(pred_classes$return_type, pred_classes$.pred_class)
```


**SPARK feature importance**

![](https://raw.githubusercontent.com/brndngrhm/big-data-bowl/main/img/feature_importance.svg)

## Limitations

The kickoff strategy of a return team is a complex decision, likely planned days if not weeks in advance, and there is undoubtedly some important predictors missing in the data. With more time and access to more data, model performance may be improved.

Additionally, the only input *not* known at time of kickoff is kick length and kick direction. I believe these variables are somewhat controllable by the kicking team/kicker, but if that is not the case then the model should not have used those as predictors.

## Conclusions

The model developed here can predict if the receiving team will return a kickoff out of their own endzone with 89% accuracy. Hopefully teams can use this to achieve some marginal benefit in their kickoff strategy. (Also I want to thank Kaggle for providing the data to let me dive into this question which I've wondered about for years!) 